National demographic-disturbance relationships and local monitoring both reduce uncertainty in projections of boreal caribou population viability: tools to support monitoring decisions

### 2023-08-28 analysis/paper/paper.Rmd
ongoing writing

---
title: "National demographic-disturbance relationships and local monitoring both reduce uncertainty in projections of boreal caribou population viability: tools to support monitoring decisions"
author:
  - name: Josie Hughes
    email: josie.hughes@ec.gc.ca
    institute: [ECCC]
    correspondence: true
  - name: Sarah Endicott
    email: sarah.endicott@ec.gc.ca
    institute: [ECCC]
    correspondence: false
  - name: Cheryl A Johnson
    email: cheryl-ann.johnson@ec.gc.ca
    institute: [ECCC]
    correspondence: false
  - name: Anna M Calvert
    email: anna.calvert@ec.gc.ca
    institute: [ECCC]
    correspondence: false
institute:
  - ECCC: Environment and Climate Change Canada 
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      number_sections: no
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: CaribouDemographicModel.bib
csl: "../templates/journal-of-forest-ecology-and-management.csl" # Insert path for the bib-style
abstract: |
  Across fifty-eight boreal caribou ranges in Canada, both survival and recruitment decrease with the percentage of the range that is disturbed. However, there is also variation in demographic rates among ranges, particularly where anthropogenic disturbance is low, and this variation leads to uncertainty in projections derived from national disturbance-demographic relationships. Demographic projections derived from local range-specific data are also uncertain because few places have enough long-term monitoring data to characterize local demographic-disturbance relationships. We present a simple Bayesian integrated population model that integrates prior information from national analysis of demographic-disturbance relationships with available local demographic data to reduce uncertainty in population viability projections. We combine this model with methods for simulating local population dynamics and monitoring to show how monitoring requirements depend on landscape condition. Where anthropogenic disturbance is very high or very low, reasonably accurate status projections can be made using only national demographic-disturbance relationships. At intermediate disturbance levels where uncertainty is higher, deploying few collars for more years may be a lower cost option for improving the reliability of status projections relative to more intensive short-term monitoring. We provide an R Shiny interface to enable others to explore the implications of alternative monitoring scenarios, disturbance trajectories, priors, and other model parameters for boreal caribou demographic projections. This tool can help users assess monitoring needs and strategies.

@@

For this investigation we assume collars are deployed in January, and last for up to 6 years. The target number of collared females $n$ varies among scenarios, as does the total duration of the monitoring program (Table \@ref(tab:monitoringScenarios)). Collars lost to mortality may be replaced each year ($o=1$), or there may be a 3 year gap between deployments ($o=4$). We focus on exploring a set of scenarios in which anthropogenic disturbance, measured as the percentage of area within 500 m of disturbance [@johnson_science_2020;@ECCC_scientific_2011], increases over time from a variety of starting points (Table \@ref(tab:monitoringScenarios)). Scenarios in which anthropogenic disturbance does not increase over time are included in Supplement D.

@@

>>>>

---
title: "National demographic-disturbance relationships and local monitoring both reduce uncertainty in projections of boreal caribou population viability: tools to support monitoring decisions"
author:
  - name: Josie Hughes
    email: josie.hughes@ec.gc.ca
    institute: [ECCC]
    correspondence: true
  - name: Sarah Endicott
    email: sarah.endicott@ec.gc.ca
    institute: [ECCC]
    correspondence: false
  - name: Cheryl A Johnson
    email: cheryl-ann.johnson@ec.gc.ca
    institute: [ECCC]
    correspondence: false
  - name: Anna M Calvert
    email: anna.calvert@ec.gc.ca
    institute: [ECCC]
    correspondence: false
institute:
  - ECCC: Environment and Climate Change Canada 
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      number_sections: no
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: CaribouDemographicModel.bib
csl: "../templates/journal-of-forest-ecology-and-management.csl" # Insert path for the bib-style
abstract: |
  Across fifty-eight boreal caribou ranges in Canada, both survival and recruitment decrease with the percentage of the range that is disturbed. However, there is also variation in demographic rates among ranges, particularly where anthropogenic disturbance is low, and this variation leads to uncertainty in projections derived from national disturbance-demographic relationships. Demographic projections derived from local range-specific data are also uncertain because few places have enough long-term monitoring data to characterize local demographic-disturbance relationships. We present a simple Bayesian population model that integrates prior information from national analysis of demographic-disturbance relationships with available local demographic data to reduce uncertainty in population viability projections. We combine this model with methods for simulating local population dynamics and monitoring to show how monitoring requirements depend on landscape condition. Where anthropogenic disturbance is very high or very low, reasonably accurate status projections can be made using only national demographic-disturbance relationships. At intermediate disturbance levels where uncertainty is higher, deploying few collars for more years may be a lower cost option for improving the reliability of status projections relative to more intensive short-term monitoring. We provide an R Shiny interface to enable others to explore the implications of alternative monitoring scenarios, disturbance trajectories, priors, and other model parameters for boreal caribou demographic projections. This tool can help users assess monitoring needs and strategies.

@@

For this investigation we assume collars are deployed in January, and last for up to 6 years. The target number of collared females $n$ varies among scenarios, as does the total duration of the monitoring program (Table \@ref(tab:monitoringScenarios)). Collars lost to mortality may be replaced each year ($o=1$), or there may be a 3 year gap between deployments ($o=4$). We focus on exploring a set of scenarios in which anthropogenic disturbance, measured as the percentage of area within 500 m of disturbance [@johnson_science_2020;@eccc_scientific_2011], increases over time from a variety of starting points (Table \@ref(tab:monitoringScenarios)). Scenarios in which anthropogenic disturbance does not increase over time are included in Supplement D.

@@

Re paragraph on costs of monitoring. The reality is that costs vary with remoteness and the difficulty of finding animals to collar (which also depends on how many collars are deployed). Since costs are so context-specific, there isn't a good way to incorporate them into this general analysis. It would be interesting to work through a case study (i.e. Newfoundland) informed by context-specific costs, and consider how incorporation of those costs may change the relative value of alternative monitoring scenarios.

Biases in composition survey estimates also vary with vegetation cover (i.e. the difficulty of seeing animals) and the care taken to correctly identify the sex of unknown animals. Misidentification rates could be estimated from discrepancies among results from multiple independent observers analyzing photos of animal groups.

The tool we have developed is a proof of concept, designed to demonstrate possibilities and help users develop a general understanding of the relative value of alternative monitoring strategies for boreal caribou. In current form the tool will have limited value for managers who have some local data and wish to know the value of collecting more, because it is designed to either integrate real local data or simulated local data, but not both. Development of a tool designed to help decision makers assess the relative value of strategies for additional monitoring given their existing local data and national-demographic disturbance relationships would be a useful followup to this work.

A tool that integrates real local data with national demographic disturbance relationships can also be used to provide consistent and transparent assessments of the strength of evidence for local deviations from the expected mean of national demographic-disturbance relationships. Help distinguish real variation among populations from apparent variation that can be attributed to lack of local data.
